Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of TESS

### 2023-07-11 paper/ms.tex
update ms

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are low-mass
  pre-main-sequence stars with nearly periodic optical modulation.
  The modulation is likely induced by dust or gas clumps in orbit at
  the Keplerian corotation radius.  Here, we report new CQVs
  discovered in TESS data collected between July~2018 and Sep~2022.
  Our search of \nstarssearched\ K- and M-dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ gold-standard CQVs.  Most of these
  discoveries are new, and they include the brightest
  ($T$$\approx$9.5) and closest ($d$$\approx$20\,pc)
  examples of this class of object known, as well as the most massive
  ($\approx$0.65\,M$_\odot$).  A few objects are outliers among
  outliers; LP 12-502 for instance shows a ``dip complex'' with a
  period and total duration that are fixed over more than 1{,}000
  cycles, but which in detail exhibits anywhere from four to eight
  local minima per cycle.  LP-502 also displayed distinct superposed
  periods simultaneously, and displayed drastic shape changes shortly
  after flares.  Broadly speaking, we find that none of the CQVs
  maintain a fixed light curve shape over timescales of more than a
  few hundred cycles, and we revisit the arguments for why transient
  corotating material is the most viable explanation.  In the future,
  we expect that our sample will facilitate modeling and observational
  efforts aimed at understanding these objects, and connecting them to
  the broader contexts of star, disk, and exoplanet evolution.  For
  instance, if the transiting material is dust, then the rate of dip
  evolution implies that of order an asteroid belt's worth of mass
  passes through these structures over the $10^8$ years for which they
  are observed.
\end{abstract}

@@

A few competing explanations for what causes the complex quasiperiodic
variability are shown in Figure~\ref{fig:f1}, along with some
representative light curves.  All young M-dwarfs are spotted, which
produces flux variations over characteristic timescales of the
rotation period, $P_{\rm rot}$, and its half-harmonic, $0.5\,P_{\rm
rot}$.  The observed dips occur over durations as short as
$0.05\,P_{\rm rot}$; a ``starspot-only'' scenario can be discarded for
any object with sufficiently sharp dips
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The timescales and
amplitudes of the flux variations instead require sharp geometries
with material extrinsic to the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  The
``clump'' scenario invokes opaque dust clumps orbiting near the
Keplerian corotation radius, which periodically eclipse the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The ``prominence''
scenario invokes long-lived condensations of cool and dense
marginally-ionized gas, embedded within the hotter corona, that would
be centrifugally supported near corotation
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
Such structures are analogous to quiescent prominences and filaments
seen in the solar corona \citep[see e.g.][]{2015ASSL..415.....V},
though at much larger relative distances from the stellar surface.  A
final possibility that has been suggested is one of a ``screen'', in
which the inner wall of a quiescent circumstellar disk blocks a
portion of the stellar surface to produce sudden dips whenever spots
come into view \citep{2019ApJ...876..127Z}.

The dust clump and prominence hypotheses seem most plausible.  They
are qualitatively similar, except that one invokes opacity from dust,
while the other invokes opacity from gas, likely bound-free
transitions in hydrogen or perhaps a molecular opacity.  The screen
scenario seems inconsistent with the degree of observed periodicity,
the lack of infrared excess, and the observed lifetime of CQVs
extending an order of magnitude longer than the $\approx$$10^7$ year
timescale typically quoted for the primordial disk to disperse.
However, unambiguous evidence has yet to be acquired in support of any
scenario.  Such evidence might include a spectroscopic detection of
silicate 10\,$\mu$m dust absorption during a dip, or perhaps detection
of transient Balmer-line absorption during the dips, similar to
observations made in systems such as AB~Dor \citep[see the review
by][]{1999ASPC..158..146C}.

CQVs remain mysterious because they have been both hard to discover
and hard to characterize.   Discoverability is tied to rarity: CQVs
comprise about 1\% of the youngest 1\% of M-dwarfs
\citep{2018AJ....155..196R}.  Out of the millions of stars monitored
by K2 and TESS, about 50 CQVs have been reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2022AJ....163..144G,2023ApJ...945..114P}.
The known CQVs are correspondingly faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, yielding optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  This renders time-series spectroscopy
at high resolution technically impossible, despite its potential
utility in ruling between the models.

One way to help rule between the mechanisms is to find bright and
nearby CQVs, since these objects will be the most amenable to detailed
photometric and spectroscopic analyses.  To do this, in this work we
use 120-second cadence data acquired by TESS between July~2018 and
Sep~2022 (Sectors 1-55; Cycles 1-4).  We present our search methods in
Section~\ref{sec:methods}, and the properties of the resulting CQV
catalog in Section~\ref{sec:results}.  The evolution of many CQVs over
a two-year baseline is described in Section~\ref{sec:evoln}, including
a deep-dive into LP 12-502.  Implications are discussed in
Section~\ref{sec:discussion}, and we conclude in
Section~\ref{sec:conclusion}.

A word on nomenclature.  CQVs have been called ``transient and
persistent flux dips'', ``scallop shells'', ``batwings'',
\citep{2017AJ....153..152S} ``complex rotators'',
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P}
and ``complex periodic variables'' \citep{2023MNRAS.518.2921K}.  The
CQVs should not be conflated with ``dippers'', which are classical T
Tauri stars with infrared excesses, and large-amplitude variability
linked to obscuring inner disk structures and accretion hot spots
\citep{2014AJ....147...82C,2021ApJ...908...16R}.  With that said, a
few similarities between CQVs and dippers do exist (see
Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}).  At the
risk of introducing yet another standard, we hope to introduce a
nomenclature that reflects how, when observed over timescales of more
than tens of cycles, CQVs are almost but not exactly periodic.  They
are quasiperiodic.  While the three-type classification scheme
proposed by \citet{2017AJ....153..152S} may indeed provide some
helpful visual distinctions amongst CQV sub-classes, it seems likely
that they are all explained by a single underlying phenomenon, and so
we opt to refer to them by a single empirically descriptive name.

@@

We analyzed the ``short'' 120-second cadence data acquired by TESS
between July~2018 and Sep~2022 (Sectors 1-55).  Specifically, we used
the 120-second cadence light curve products produced by the mission's
Science Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products also
include full frame images with cadences of 200, 600, and 1800 seconds,
we limited our scope in this work for the sake of simplicity in data
handling.  In exchange, we lose in both completeness and homogeneity
of the selection function.  While TESS cumulatively observed
$\approx$90\% of the sky for at least one lunar month between
July~2018 and Sep~2022, the 120-second cadence data were acquired for
only a pre-selected set of stars over Sectors 1-26, and then a
guest-investigator driven set of stars over Sectors 27-55
\citep{2021PASP..133i5002F}.

@@

Table~\ref{tab:thetable} lists the \ncpvsfound\ objects identified by
our search.  \ngoods\ objects demonstrated what we viewed as
unambiguous characteristics of the CQV phenomenon in at least one TESS
sector; we refer to these as the ``gold sample''.  We found the
remaining \nmaybes\ CQV candidates to be ambiguous; their variability
might be caused by say, multiple starspot groups.  Additional data
from TESS or other instruments could help resolve their
classification.  The boolean \texttt{quality} column in the table divides
the two classes.  In the following discussion, we will restrict our
discussion to the gold sample.

@@

We can estimate the mass of a transiting cloud by first converting the
transit depth to an effective cloud radius, $R_{\rm cloud}$.  For most
CQVs in Figure~\ref{fig:cqvs}, this typically yields $\approx$2-20\,$R_\oplus$.
This size can be converted into a mass estimate by requiring the
optical depth $\tau$ of the cloud to be at least unity and by positing
some composition for the cloud.  In other
words, a minimum constraint on the number density follows by
requiring the cloud to be optically thick.  For cases like LP 12-502,
this is resonable because the transit duration implies $R_{\rm
cloud}\ll R_\star$.  Carrying out the relevant calculation assuming
the occulting material is dust grains 1\,$\mu$m in size,
\citet{2023MNRAS.518.4734S} find minimum cloud masses of order
$10^{12}$\,kg (their Eq.~23), which scale linearly with both the
optical depth and dust grain radius.  This is roughly comparable to a
small asteroid; the asteroid belt itself has a mass of
order $\approx$10$^{21}$\,kg (CITE).  Also by way of comparison,
the previously-discussed cool gas prominences for AB~Dor would need to
have masses of order $10^{14}$\,kg \citep{1990MNRAS.247..415C}, about
100$\times$ larger than the requisite dust mass.

Our observations provide a direct measurement of how often dips
appear and disappear, both due to sudden state-switches, and
due to more gradual, secular evolution.  For instance, LP~12-502
showed three ``state-switch'' events over the six months of available TESS
observations, during cycles 248-264, 299-315, and 1233-1247.  It is
plausible to imagine that these events correspond to either mass being
ejected from corotation, or perhaps being accreted onto the star.  In
either case, the corresponding $\dot{M} \equiv M\cdot {\rm d}N/{\rm d}t$
is $\approx$3$\times$$10^{-18} M_\odot\,{\rm yr}^{-1} \approx
1\times10^{-12} M_\oplus\,{\rm yr}^{-1}$.  Considered cumulatively
over the $\approx$$10^8$ years for which the CQV phenomenon is
observed, this yields a cumulative moved dust mass of
$10^{-4}\,M_\oplus$, of order the Solar System's asteroid belt.
If the occulting material is gas, the masses involved would be of
order 100 times larger.

>>>>

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are low-mass
  pre-main-sequence stars with nearly periodic optical modulation.
  The modulation is likely induced by dust or gas clumps in orbit at
  the Keplerian corotation radius.  Here, we report new CQVs
  discovered in TESS data collected between July~2018 and Sep~2022.
  Our search of \nstarssearched\ K- and M-dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CQVs.  Most of these
  discoveries are new, and they include the brightest
  ($T$$\approx$9.5) and closest ($d$$\approx$20\,pc)
  examples of this class of object known, as well as the most massive
  ($\approx$0.6-0.8\,M$_\odot$).  A few objects are outliers among
  outliers; LP 12-502 for instance shows a ``dip complex'' with a
  period and total duration that are fixed over more than 1{,}000
  cycles, but which in detail exhibits anywhere from four to eight
  local minima per cycle.  LP-502 also displayed distinct superposed
  periods simultaneously, and showed drastic shape changes shortly
  after flares.  Broadly speaking, we find that none of the CQVs
  maintain a fixed light curve shape over timescales of more than a
  few hundred cycles, and we revisit the arguments for why transient
  corotating material is the most viable explanation.  In the future,
  we expect that our sample will facilitate modeling and observational
  efforts aimed at understanding these objects, and connecting them to
  the broader contexts of star, disk, and exoplanet evolution.  For
  instance, if the transiting material is dust, then the rate of dip
  evolution implies that of order an asteroid belt's worth of mass
  passes through these structures over their lifetime of $10^8$ years.
\end{abstract}

@@

A few competing explanations for what causes the complex quasiperiodic
variability are shown in Figure~\ref{fig:f1}.  All young M-dwarfs are spotted, which
produces flux variations over characteristic timescales of the
rotation period, $P_{\rm rot}$, and its half-harmonic, $0.5\,P_{\rm
rot}$.  The observed dips occur over durations as short as
$0.05\,P_{\rm rot}$; a ``starspot-only'' scenario can be discarded for
any object with sufficiently sharp dips
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The timescales and
amplitudes of the flux variations instead require sharp geometries
with material extrinsic to the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  The
``clump'' scenario invokes opaque dust clumps orbiting near the
Keplerian corotation radius, which periodically eclipse the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The ``prominence''
scenario invokes long-lived condensations of cool and dense
marginally-ionized gas, embedded within the hotter corona, that would
be centrifugally supported near corotation
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
Such structures are analogous to quiescent prominences and filaments
seen in the solar corona \citep[see e.g.][]{2015ASSL..415.....V},
though at much larger relative distances from the stellar surface.  A
final possibility that has been suggested is one of a ``screen'', in
which the inner wall of a quiescent circumstellar disk blocks a
portion of the stellar surface to produce sudden dips whenever spots
come into view \citep{2019ApJ...876..127Z}.

The dust clump and prominence hypotheses seem most plausible.  They
are qualitatively similar, except that one invokes opacity from dust,
while the other invokes opacity from gas, likely bound-free
transitions in hydrogen or perhaps a molecular opacity.  The screen
scenario seems inconsistent with the degree of observed periodicity,
the lack of infrared excess, and the observed lifetime of CQVs
extending an order of magnitude longer than the $\approx$$10^7$ year
timescale typically quoted for primordial disk dispersal.
However, unambiguous evidence in support of any one scenario has yet
to be acquired.  Such evidence might include a spectroscopic detection
of silicate 10\,$\mu$m dust absorption during a dip, or perhaps
detection of transient Balmer-line absorption as a function of cycle
phase, similar to observations made in systems such as AB~Dor
\citep[see the review by][]{1999ASPC..158..146C}.

CQVs remain mysterious because they have been both hard to discover
and hard to characterize.   Discoverability is tied to rarity: CQVs
comprise $\approx$1\% of the youngest $\approx$1\% of M-dwarfs
\citep{2018AJ....155..196R}.  Out of the millions of stars monitored
by K2 and TESS, about 50 CQVs have been reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2022AJ....163..144G,2023ApJ...945..114P}.
The known CQVs are correspondingly faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, yielding optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  This renders time-series spectroscopy
at high resolution out of reach for current facilities, despite its
potential utility in ruling between the models.

In this work, we aim to find bright and nearby CQVs, since these
objects will be the most amenable to detailed photometric and
spectroscopic analyses.  To do this, we use 120-second cadence data
acquired by TESS between July~2018 and Sep~2022 (Sectors 1-55; Cycles
1-4).  We present our search methods in Section~\ref{sec:methods}, and
the properties of the resulting CQV catalog in
Section~\ref{sec:results}.  The evolution of many CQVs over a two-year
baseline is described in Section~\ref{sec:evoln}, including a
deep-dive into LP 12-502.  Implications are discussed in
Section~\ref{sec:discussion}, and we conclude in
Section~\ref{sec:conclusion}.

A point on nomenclature.  CQVs have been called ``transient flux
dips'', ``persistent flux dips'', ``scallop shells'', ``batwings'',
\citep{2017AJ....153..152S} ``complex rotators'',
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P}
and ``complex periodic variables'' \citep{2023MNRAS.518.2921K}.  The
CQVs should not be conflated with ``dippers'', which are classical T
Tauri stars with infrared excesses, and large-amplitude variability
linked to obscuring inner disk structures and accretion hot spots
\citep{2014AJ....147...82C,2021ApJ...908...16R}.  With that said, a
few similarities between CQVs and dippers do exist (see
Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}).  At the
risk of introducing yet another standard, we hope to introduce a
nomenclature that reflects how, when observed over timescales of more
than tens of cycles, CQVs are almost but not exactly periodic.  They
are quasiperiodic.  While the three-type classification scheme
proposed by \citet{2017AJ....153..152S} may indeed provide some
helpful visual distinctions amongst the CQVs, it seems likely
that they are all explained by a single underlying phenomenon, and so
we opt to refer to them by a single empirically descriptive name.

@@

We analyzed the ``short'' 120-second cadence data acquired by TESS
between July~2018 and Sep~2022 (Sectors 1-55).  Specifically, we used
the 120-second cadence light curve products produced by the 
Science Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products also
include full frame images with cadences of 200, 600, and 1800 seconds,
we limited our scope in this work for the sake of simplicity in data
handling.  In exchange, we lose in both completeness and homogeneity
of the selection function.  While TESS cumulatively observed
$\approx$90\% of the sky for at least one lunar month between
July~2018 and Sep~2022, the 120-second cadence data were acquired for
only a pre-selected set of stars over Sectors 1-26, and then a
guest-investigator driven set of stars over Sectors 27-55
\citep{2021PASP..133i5002F}.

@@

Table~\ref{tab:thetable} lists the \ncpvsfound\ objects identified by
our search.  \ngoods\ objects demonstrated what we viewed as
unambiguous characteristics of the CQV phenomenon in at least one TESS
sector; we refer to these as the ``high-quality sample''.  We found the
remaining \nmaybes\ CQV candidates to be ambiguous; their variability
might be caused by say, multiple starspot groups.  Additional data
from TESS or other instruments could help resolve their
classification.  The boolean \texttt{quality} column in the table divides
the two classes.  In the following discussion, we will restrict our
discussion to the high-quality sample.

@@

We can estimate the mass of a transiting cloud by first converting the
transit depth to an effective cloud radius, $R_{\rm cloud}$.  For most
CQVs in Figure~\ref{fig:cqvs}, this typically yields $\approx$2-20\,$R_\oplus$.
This size can be converted into a mass estimate by requiring the
optical depth $\tau$ of the cloud to be at least unity and by positing
some composition for the cloud.  In other
words, a minimum constraint on the number density follows by
requiring the cloud to be optically thick.  For cases like LP 12-502,
this is reasonable because the transit duration implies $R_{\rm
cloud}\ll R_\star$.  Carrying out the relevant calculation assuming
the occulting material is dust grains 1\,$\mu$m in size,
\citet{2023MNRAS.518.4734S} find minimum cloud masses of order
$10^{12}$\,kg (their Eq.~23), which scale linearly with both the
optical depth and dust grain radius.  This is comparable to a
small asteroid; the asteroid belt itself has a mass of
order $\approx$10$^{21}$\,kg (CITE).  
The previously discussed cool gas prominences would need to have
masses of order $10^{14}$\,kg \citep{1990MNRAS.247..415C}, about
100$\times$ larger than the requisite dust mass.

Our observations provide a direct measurement of how often dips
appear and disappear, both due to sudden state-switches, and
due to more gradual, secular evolution.  For instance, LP~12-502
showed three ``state-switch'' events over the six months of available TESS
observations, during cycles 248-264, 299-315, and 1233-1247.  In each
case, a dip ``turned off''.  It is
plausible to imagine that these events correspond to either mass being
ejected from corotation, or perhaps being accreted onto the star.  In
either case, the corresponding $\dot{M} \equiv M\cdot {\rm d}N/{\rm d}t$
time-averaged over six months
is $\approx$3$\times$$10^{-18} M_\odot\,{\rm yr}^{-1} \approx
1\times10^{-12} M_\oplus\,{\rm yr}^{-1}$.  Considered cumulatively
over the $\approx$$10^8$ years for which the CQV phenomenon is
observed, this yields a cumulative moved dust mass of
$10^{-4}\,M_\oplus$, of order the Solar System's asteroid belt.
If the occulting material is gas, the masses involved would be of
order 100 times larger.
For cases in which we observe the {\it growth} of
dips, such as the Sector~29 data for TIC~224283342, the dip depths
typically increase by of order a few percent over ten to twenty days.
{\bf fixme: add closing sentence}
